A comparative study of batch construction strategies \\ for recurrent neural networks in MXNet

### 2017-04-19 paper/paper.tex
semi-final, 3 WERs missing

  \maketitle
  %
  \begin{abstract}
  	 In this work we compare different methods for sequence selection for mini-batch training
  	 of recurrent neural networks. While popular implementations like TensorFlow and MXNet suggest 
  	 a bucketing approach to improve the parallelization capabilities of the recurrent training process, 
  	 we propose several simple ordering strategies that are able to compete both in training time and 
  	 recognition performance while being conceptually simpler. We compare our method with various other
  	 batch construction strategies on the CHiME-4 noisy speech recognition corpus. The experiments were
  	 done in MXNet and a demo implementation will be provided for free use.
  \end{abstract}
  \noindent{\bf Index Terms}: MXNet, bucketing, batches, recurrent neural networks

  \section{Introduction}
  	Neural network based acoustic modeling became the de-facto standard in automatic speech recognition (ASR)
  	and related tasks. Modeling contextual information over long distances in the input signal hereby showed to 
  	be of fundamental importance for optimal system performance. Modern acoustic models therefore use recurrent 
  	neural networks (RNN) to model long temporal dependencies. In particular the long short-term memory (LSTM)
  	has been shown to work very well on these tasks and most current state-of-the-art systems incorporate LSTMs
  	in their acoustic models. While it is common practice to train the
  	models on a frame-by-frame labeling obtained from a 
  	previously trained system, sequence-level criteria that optimize the acoustic model and the alignment model jointly
  	are becoming increasingly popular. As an example, the connectionist temporal classification (CTC) \cite{CTC}
  	enables fully integrated training of acoustic models without assuming a frame-level alignment to be given. Sequence-level
  	criteria however require to train on full utterances, where it is possible to train frame-wise labeled systems
  	on sub-sequences of any resolution.

@@

    In this work we propose a simple batch construction strategy that does not rely on such a hyper-parametrization. The method 
    produces batches with large variability of sequences while at the same time reducing irrelevant computation to a minimum.
    The following sections we are going to give an overview over current batch construction strategies and compare them 
    w.r.t.~training time and variability. We will then derive our proposed method and discuss its properties on a theoretical
    level, followed by an empirical evaluation on the CHiME-4 noisy speech recognition task. All experiments were implemented 
    within the MXNet framework and are available for download on our website.

@@

  Bucketing is especially suited if the RNN model itself is not able to handle arbitrary 
  long sequences but instead requires to store an unrolled version of the network for every possible 
  length. In those cases bucketing allows the framework to assign each batch to the shortest possible 
  unrolled network, while still optimizing the same shared weights.

@@

@@

 \begin{minipage}{\textwidth}
 	For each epoch
 	\begin{enumerate}
 		\item shuffle training data by sequence length
 		\item partition resulting sequence order into $N$ partitions
 		\item sort each bin $n$:
 		\begin{itemize}
 			\item in increasing order if $n$ is odd
 			\item in decreasing order if $n$ is even
 		\end{itemize}
 		\item partition into batches of desired size \\[1ex]
 	\end{enumerate}
 \end{minipage}

 Due to the initial shuffling and subsequent partitioning the probability for two sequences of any length being 
 put into the same bin in $\frac{1}{N*(N-1)}$, so by increasing the number of bins, the variability within a 
 partition decreases quadratically while the variability among different partitions increases. The alternated
 sorting approach ensures that utterances at the boundaries of two consecutive bins are of similar length such 
 that the final partitioning into batches requires minimal zero padding.

@@

  \section{Experiments} \label{sec:experiments}
   In order to provide some perspective on the required context on the CHiME-4 task, 
   we provide results for experiments on sub-utterance (chunk) in 
   Table \ref{tab:chime:chunk}. In the first columns we evaluate different sub-sequence lengths, while measuring the utterances that were processed per second, the memory required and the word error rate (WER) on the evaluation set of the CHiME-4 database. We can observe that 
   while large speed-ups can be obtained when training is done in this fashion, full-utterance context is
   required for optimal performance.

@@

	\begin{table}[tbp]
		\centering
		\caption{Training time and recognition performance when training is done 
		on sub-sequence level. The first column shows the maximal sub-sequence length 
		after partitioning of the original sequence. The last line shows the results 
		if no partitioning into sub-sequence is performed. }
		\label{tab:chime:chunk}
		\begin{tabular}{lrrr}
			\hline
			Chunk size              & Utt./sec & Memory [GB] & WER [\%] \\
			\hline
			10     					&				 &			   &			 \\
			50  					&				 &			   &    		 \\
			100 					& 				 &			   &			 \\
			500						& 				 &			   &			 \\
			$\max$					&				 &			   &			 \\
			\hline
		\end{tabular}
	\end{table}

	\begin{table}[tbp]
		\centering
		\caption{An evaluation of different sequence batch construction methods on the CHiME-4 database. Training time per epoch, memory consumption are presented in the first two columns, while the last columns shows the word error rate of the corresponding system
		when decoded with the RASR speech recognition toolkit.}
		\label{tab:chime:batch}
		\begin{tabular}{lrrr}
			\hline
			Approach                & Utt./sec & Memory [GB] & WER [\%]          \\
			\hline
			Random 					&	       & 			 & 6.8		 \\
			Sorted					& 		   &			 & 7.4		 \\
			\hline
			Bucketing				&		   &			 & 7.2		 \\
			Proposed (4 bins)	    &		   &			 & 7.0		 \\
			Proposed (64 bins)	    &		   &			 & 7.0		 \\
			Proposed (256 bins)	    &		   &			 & 6.9		 \\
			\hline
		\end{tabular}
	\end{table}

  As expected, sorting the entire training set by sequence length reduces the required time 
  per epoch to a minimum, while the best overall performance is obtained when sequences are shuffled randomly.

@@

   apptek?

>>>>

  \maketitle
  %
  \begin{abstract}
  	 In this work we compare different methods for sequence selection for mini-batch training
  	 of recurrent neural networks. While popular implementations like TensorFlow and MXNet suggest 
  	 a bucketing approach to improve the parallelization capabilities of the recurrent training process, 
  	 we propose several simple ordering strategies that are able to compete both in training time and 
  	 recognition performance while being conceptually simpler. We compare our method with various other
  	 batch construction strategies on the CHiME-4 noisy speech recognition corpus. 
  \end{abstract}
  \noindent{\bf Index Terms}: MXNet, bucketing, batches, recurrent neural networks

  \section{Introduction}
  	Neural network based acoustic modeling became the de-facto standard in automatic speech recognition (ASR)
  	and related tasks. Modeling contextual information over long distances in the input signal hereby showed to 
  	be of fundamental importance for optimal system performance. Modern acoustic models therefore use recurrent 
  	neural networks (RNN) to model long temporal dependencies. In particular the long short-term memory (LSTM)~\cite{hochreiter1997lstm}
  	has been shown to work very well on these tasks and most current state-of-the-art systems incorporate LSTMs
  	in their acoustic models. While it is common practice to train the
  	models on a frame-by-frame labeling obtained from a 
  	previously trained system, sequence-level criteria that optimize the acoustic model and the alignment model jointly
  	are becoming increasingly popular. As an example, the connectionist temporal classification (CTC) \cite{CTC}
  	enables fully integrated training of acoustic models without assuming a frame-level alignment to be given. Sequence-level
  	criteria however require to train on full utterances, where it is possible to train frame-wise labeled systems
  	on sub-sequences of any resolution.

@@

    In this work we propose a simple batch construction strategy that does not rely on such a hyper-parametrization. The method 
    produces batches with large variability of sequences while at the same time reducing irrelevant computation to a minimum.
    The following sections we are going to give an overview over current batch construction strategies and compare them 
    w.r.t.~training time and variability. We will then derive our proposed method and discuss its properties on a theoretical
    level, followed by an empirical evaluation on the CHiME-4 noisy speech recognition task. 
    %All experiments were implemented 
    %within the MXNet framework and are available for download on our website.

@@

  Bucketing is especially useful if the RNN model itself does not support dynamic unrolling and 
  is not able to handle arbitrary 
  long sequences but instead requires to store an unrolled version of the network for every possible 
  length. In those cases bucketing allows the framework to assign each batch to the shortest possible 
  unrolled network, while still optimizing the same shared weights.

@@

  \begin{figure}
   	\resizebox{0.45\textwidth}{!}{\includegraphics{plot}}
   	\caption{Resulting sequence ordering for different batch construction strategies. The graphs 
   	show the length distribution for 1000 randomly selected samples of the CHiME-4 training set. On the 
   	vertical axis the length is plotted and on the X axis the utterance order. Bucketing was done with a bucket size of 250 and in the proposed approach we used 12 bins.}
   	  \label{fig:plot}
   \end{figure}

@@

 \begin{minipage}{\textwidth}
 	For each epoch
 	\begin{enumerate}
 		\item shuffle training data by sequence length
 		\item partition resulting sequence order into $N$ partitions
 		\item sort each bin $n$:
 		\begin{itemize}
 			\item in increasing order if $n$ is odd
 			\item in decreasing order if $n$ is even
 		\end{itemize}
 		\item partition resulting order into batches of desired size \\[1ex]
 	\end{enumerate}
 \end{minipage}

 Due to the initial shuffling and subsequent partitioning the probability for two sequences of any length being 
 put into the same bin in $\frac{1}{N\cdot(N-1)}$, so by increasing the number of bins, the variability within a 
 partition decreases quadratically while the variability among different partitions increases. The alternated
 sorting approach ensures that utterances at the boundaries of two consecutive bins are of similar length such 
 that the final partitioning into batches requires minimal zero padding.

 Figure \ref{fig:plot} shows the length distribution for random and sorted sequence ordering as well as for bucketing in MXNet and the proposed approach. Note that in the case of bucketing batches are put
 together by randomly choosing one of the buckets first, so the ordering does not directly represent 
 the final set of batches.

@@

  \section{Experiments} \label{sec:experiments}
   In order to provide some perspective on the required context on the CHiME-4 task, 
   we provide results for experiments on sub-utterance (chunk) in 
   Table \ref{tab:chime:chunk}. In the first columns we evaluate different sub-sequence lengths, while measuring the utterances that were processed per second, the memory required and the word error rate (WER) on the evaluation set of the CHiME-4 database. We hereby constrained batches to only 
   contain 5,000 frames in total, such that the same number of updates is constant in all experiments. We can observe that 
   while large speed-ups can be obtained when training is done in this fashion, full-utterance context is
   required for optimal performance. However, it is worth noting that the memory requirement decreases 
   significantly when sub-utterance training is applied and especially for unusually long utterances, 
   sub-utterance training might be the only way to fit the data into GPU memory.

@@

	\begin{table}[tbp]
		\centering
		\caption{Training time and recognition performance when training is done 
		on sub-sequence level. The first column shows the maximal sub-sequence length 
		after partitioning of the original sequence. The last line shows the results 
		if no partitioning into sub-sequence is performed. }
		\label{tab:chime:chunk}
		\begin{tabular}{lrrr}
			\hline
			Chunk size              & Utt./sec & Memory [GB] & WER [\%] \\
			\hline
			10     					&	36.7   & 1.6         &			 \\
			50  					&	31.1   & 1.6		 &    		 \\
			100 					& 	29.6   & 1.6		 &			 \\
			500						& 	17.3   & 1.6         &			 \\
			$\max$					&	7.0	   & 5.4         &	8.9		 \\
			\hline
		\end{tabular}
	\end{table}

	\begin{table}[tbp]
		\centering
		\caption{An evaluation of different sequence batch construction methods on the CHiME-4 database. Training time per epoch, memory consumption are presented in the first two columns, while the last columns shows the word error rate of the corresponding system
		when decoded with the RASR speech recognition toolkit.}
		\label{tab:chime:batch}
		\begin{tabular}{lrrr}
			\hline
			Approach                & Utt./sec & Memory [GB] & WER [\%]          \\
			\hline
			Random 					& $\sim$ 7.0     & $\sim$ 5.4		 & $\sim$ 8.9		 \\
			Sorted					& 10.2    &	4.2 & 10.2		 \\
			\hline
			Bucketing				& 9.5     &	6.3		 & 9.6		 \\
			Proposed (8 bins)	    & 10.1	   & 4.8     & 9.9		 \\
			Proposed (64 bins)	    & 10.0     & 5.3     & 9.5		 \\
			Proposed (256 bins)	    & 8.8	   & 6.0     & 9.1		 \\
			\hline
		\end{tabular}
	\end{table}

  As expected, sorting the entire training set by sequence length reduces the required time 
  per epoch to a minimum, while the best overall performance is obtained when sequences are shuffled randomly. Both bucketing and the proposed 
  approach are in between. We can observe that our method is able to reach almost the same recognition performance as using a randomly shuffled sequence
  ordering, while being almost as fast as the sorted sequence scheduler. This allows for a good trade-off between runtime and system performance.

@@

   This work has received funding from the European Research Council
   (ERC) under the European Union’s Horizon 2020 research and innovation
   program (SEQCLAS, grant agreement No 694537).  The work reflects
   only the authors' views and the European Research Council Executive
   Agency is not responsible for any use that may be made of the
   information it contains.  We further want to thank Jahn Heymann, Lukas
   Drude and Reinhold H\"ab-Umbach from University of Paderborn, Germany
   for their CHiME-4 front-end which we used in this work.
